


{{< include ../math.qmd >}}



# Probability theory

## Notations
- $Y$: a random variable (captial letters)
- $y$: a sample of $Y$
- $\Pr\qty(Y\in A\mid\theta)$: the probability of $Y$ being in $A$
- $p(y\mid\theta)=\Pr\qty(Y=y\mid\theta)$: the discrete probability density function
- $f(y\mid\theta)=\displaystyle\dv{y}\Pr\qty(Y\leq y\mid\theta)$: the continuous probability density function
- $\Exp\qty(Y)$: the expectation of $Y$
- $\Var\qty(Y)$: the variance of $Y$



## Random variables


::: {#def-}
# Expectation
$$
\Exp\mqty[u(X)] = \int_{-\infty}^{\infty}u(x)f(x)\dl3x.
$$
:::


::: {#def-}
1. $\mu=\Exp(X)$ is called the **mean value** of $X$.
2. $\sigma^2=\Var(X)=\Exp\mqty[(X-\mu)^2]$ is called the **variance** of $X$.
3. $M_X(t)=\Exp\mqty[\me^{tX}]$ is called the **moment generating function** of $X$.
:::


::: {#prp-}
1. $\Exp\mqty[ag(X)+bh(X)]=a\Exp\mqty[g(X)]+b\Exp\mqty[h(X)]$.
2. $\Var\mqty[X]=\Exp\mqty[(X-\mu)^2]=\Exp(X^2)-\mu^2$.
:::

::: {.proof}

$$
\begin{split}
\Exp\mqty[ag(X)+bh(X)]&=\int_{-\infty}^{\infty}\mqty[ag(x)+bh(x)]f(x)\dl3x\\
                 &=a\int_{-\infty}^{\infty}g(x)f(x)\dl3x+b\int_{-\infty}^{\infty}h(x)f(x)\dl3x\\
                 &=a\Exp\mqty[g(X)]+b\Exp\mqty[h(X)].
\end{split}
$$

$$
\begin{split}
\Exp\mqty[(X-\mu)^2]&=\Exp\mqty[\qty(X^2-2\mu X+\mu^2)]=\Exp(X^2)-2\mu\Exp(X)+\Exp(\mu^2)\\
&=\Exp(X^2)-2\mu\mu+\mu^2=\Exp(X^2)-\mu^2.
\end{split}
$$
:::

### R code
R has built-in random variables with different distributions. The naming convention is a prefix `d-`, `p-`, `q-` and `r-` together with the name of distribution. 

- `d-`: density function of the given distribution;
- `p-`: cumulative density function of the given distribution;
- `q-`: quantile function of the given distribution (which is the inverse of `p-` function);
- `r-`: random sampling from the given distribution.


::: {.callout-tip collapse="true"}
# Examples: normal distribution


```{r}
x <- seq(-4, 4, length=100)
y <- dnorm(x, mean=2, sd=0.5)
plot(x, y, type="l")
```


```{r}
x <- seq(-4, 4, length=100)
y <- pnorm(x, mean=2, sd=0.5)
plot(x, y, type="l")
```


```{r}
qnorm(0)
qnorm(0.5)
qnorm(1)
```


```{r}
rnorm(10)
```


:::

## Random vectors



::: {#def-randomvector}
# Random vector
Given a random experiment with a sample space $\mathcal C$, consider two random variables $X_1$ and $X_2$, which assign to each element $c$ of $\mathcal C$ one and only one ordered pair of numbers $X_1(c)=x_1$, $X_2(c)=x_2$. Then we say that $(X_1, X_2)$ is a **random vector**. The **space** of $(X_1, X_2)$ is the set of orderd pairs $\mathcal D=\set{(x_1, x_2)\mid x_1=X_1(c), x_2=X_2(c), c\in\mathcal C}$.
:::


::: {#def-cdf}
# Joint Cumulative Distribution Function
**The joint cumulative distribution function** of $(X_1, X_2)$ is defined as follows.

<!-- $$
F_{X_1,X_2}(x_1,x_2)=\pr\sqb{\set{X_1\leq x_1}\cap\set{X_2\leq x_2}}.
$$
In continuous case,  -->
$$
F_{X_1,X_2}(x_1,x_2)=\int_{-\infty}^{x_1}\int_{-\infty}^{x_2}f_{X_1,X_2}(w_1,w_2)\dl3w_1\dl3w_2.
$$
:::


<!-- ::: {#def-pdf}
# Joint probability mass function (pmf)
In discrete random vector case, the **pmf** is defined as  
$$
p_{X_1,X_2}(x_1,x_2)=\pr\sqb{X_1=x_1,X_2=x_2}.
$$
::: -->

::: {#def-cdf}
# Joint probability density function (pdf)
In continuous random vector case, the **pdf** is defined as  
$$
f_{X_1, X_2}(x_1, x_2)=\frac{\partial^2F_{X_1, X_2}(x_1,x_2)}{\partial x_1\partial x_2}.
$$
:::




::: {#def-marginal}
## Marginal pdf
Assume $(X_1, X_2)$ be a continuous random vector. The **marginal pdf** is 
$$
f_{X_1}(x_1)=\int_{-\infty}^{\infty}f(x_1, x_2)\dl3x_2.
$$
:::




::: {#def-}
# Expectation
Assume that $Y=g(X_1, X_2)$. Then
$$
\Exp(Y)=\int_{-\infty}^{\infty}\int_{-\infty}^{\infty} g(x_1, x_2) f_{X_1, X_2}(x_1,x_2)\dl3x_1\dl3x_2.
$$
:::





::: {#def-conditionaldistribution}
# Conditional probability
The conditional pdf is defined as follows:
$$
f_{X_1\mid X_2}(x_1\mid x_2)=\frac{f_{X_1,X_2}(x_1,x_2)}{f_{X_1}(x_1)}=\frac{f_{X_1,X_2}(x_1,x_2)}{\int_{-\infty}^{\infty} f_{X_1,X_2}(x_1,w)\dl3w},
$$
and the corresponding conditional probability is defined as
$$
\Pr(X_1\in A\mid X_2=x_2)=\int_Af_{X_1\mid X_2}(x_1\mid x_2)\dl3x_1.
$$
Thus $f_{X_1\mid X_2}(x_1\mid x_2)$ is a pdf of a random function of $X_1$.
:::


::: {#thm-}
Let $(X_1, X_2)$ be a random vector such that $\Var(X_2)$ is finite. $\Exp(X_2\mid X_1=x_1)$ and $\Var(X_2\mid X_1=x_1)$ can be seen as random functions of $X_1$. Then

a. $\Exp\mqty[\Exp(X_2\mid X_1)]=\Exp(X_2)$.
b. $\Var\mqty[\Var(X_2\mid X_1)]\leq\Var(X_2)$.
:::



### Relations to single variable case
Under the assumption of two variables $X$ and $Y$, when only talking about one variable $X$ (resp. $Y$), we are actually talking about the random variable corresponding to the marginal pdf. The ignoring the other variable part is handled by the integration part.


::: {#prp-}
1. $\Exp_X\mqty[u(X)]=\Exp\mqty[u(X)]$.
2. $\Var_X(X)=\Var(X)$.
:::


::: {.proof}
$$
\begin{aligned}
\Exp\mqty[u(X)]&=\iint u(x)f(x,y)\dl3x\dl3y=\int u(x) \mqty[\displaystyle\int f(x,y)\dl3y]\dl3x=\int u(x)f_X(x)\dl3x=\Exp_X\mqty[u(X)],\\
\Var(X)&=\Exp\mqty[(X-\mu)^2]=\Exp(X^2)-\mu^2=\Exp_X(X^2)-\mu^2=\Var_X(X).
\end{aligned}
$$

:::


## Maximal likelihood estimation
Consider the Bayes' Theorem 

$$
    p(\vb w\mid \mathcal D)=\frac{p(\mathcal D\mid \vb w)p(\vb w)}{p(\mathcal D)}.
$$

$p(\mathcal D\mid \vb w)$ is called the *likelihood function*, $p(\vb w)$ is called the *prior probability* and $p(\vb w\mid \mathcal D)$ is called the *posterior probability*. A widely used frequentist estimator is *maximum likelihood*, in which $\vb w$ is set to the value that maximizes the likelihood function $p(\mathcal D\mid \vb w)$. Sometimes the likelihood function is changed to be the *error function* $-\ln p$ and to maximize the likelihood function is the same as to minimize the error function.



Consider a data set of observations $\vb x=(x_1,x_2,\ldots,x_N)^T$. These data are i.i.d., with respect to the Gaussian distribution $\mathcal N(\mu,\sigma^2)$. Then we have the *likelihood function* if it is treated as a function of $\mu$ and $\sigma^2$:

$$
    p(\vb x\mid \mu,\sigma^2)=\prod_{n=1}^N\mathcal N(x_n\mid \mu,\sigma^2).
$$
We want to find $\mu$ and $\sigma^2$ to maximize the likelihood function. To do so, we would like to consider the error function 

$$
\begin{split}
    -\ln p(\vb x\mid \mu,\sigma^2)&=-\ln \prod_{n=1}^N\frac{1}{\sqrt{2\pi\sigma^2}}\exp{-\frac{1}{2\sigma^2}(x_n-\mu)^2}=\sum_{n=1}^N\qty(\frac12\ln(2\pi\sigma^2)+\frac{1}{2\sigma^2}(x_n-\mu)^2)\\
    &=\frac{1}{2\sigma^2}\sum_{n=1}^N(x_n-\mu)^2+\frac{N}{2}\ln(\sigma^2)+\frac{N}{2}\ln(2\pi).
\end{split}
$$

Take the derivative of it. We have

$$
   \begin{split}
       \pdv{ \qty(-\ln p(\vb x\mid \mu,\sigma^2))}{\mu}&=\frac{1}{2\sigma^2}\sum_{n=1}^N2(x_n-\mu)(-1)=-\frac{1}{\sigma^2}\qty(N\mu-\sum_{n=1}^Nx_n),\\
             \pdv{ \qty(-\ln p(\vb x\mid \mu,\sigma^2))}{\sigma^2}&=\frac12(-1)(\sigma^2)^{-2}\qty(\sum_{n=1}^N(x_n-\mu)^2)+\frac{N}{2}\frac{1}{\sigma^2}\\
             &=-\frac N{2(\sigma^2)^2}\qty(\frac1N\sum_{n-1}^N(x_n-\mu)^2-\sigma^2).
   \end{split} 
$$
To minimize the error function we need to let them be $0$. Then we have

$$
    \mu_{ML}=\sum_{n=1}^Nx_n,\quad \sigma^2_{ML}=\frac1N\sum_{n=1}^N(x_n-\mu_{ML})^2.
$$
<!-- These are called the *sample mean* and the *sample variance* of the data. Note that the sample variance is measured with respect to the sample mean. -->


Compute $\Exp\mqty[\mu_{ML}]$ and $\Exp\mqty[\sigma^2_{ML}]$.

$$
    \Exp\mqty[\mu_{ML}]=\Exp\mqty[\frac1N\sum_{n=1}^Nx_n]=\frac1N\sum_{n=1}^N\Exp\mqty[x_n]=\frac1NN\mu=\mu.
$$
Since $\Var\mqty[kx]=\Exp\mqty[(kx)^2]-(\Exp\mqty[kx])^2=k^2\Exp\mqty[x^2]-k^2\Exp\mqty[x]^2=k^2\Var\mqty[x]$, we have

$$
\begin{aligned}
    \Var\mqty[\mu_{ML}]&=\Var\mqty[\frac1N\sum_{n=1}^Nx_n]=\frac1{N^2}\Var\mqty[\sum_{n=1}^Nx_n]=\frac{1}{N^2}\sum_{n=1}^N\Var\mqty[x_n]\\
    &=\frac{1}{N^2}(N\sigma^2)=\frac{1}{N}\sigma^2,\\
    \Var\mqty[x_n-\mu_{ML}]&=\Var\mqty[\frac{N-1}{N}x_n-\frac1Nx_1-\ldots-\frac1Nx_N]\\
    &=\frac{(N-1)^2}{N^2}\sigma^2+\frac{1}{N^2}\sigma^2+\ldots+\frac1{N^2}\sigma^2\\
    &=\frac{N-1}{N}\sigma^2.
\end{aligned}
$$
Then 

$$
\begin{split}
    \Exp\mqty[\sigma^2_{ML}]&=\Exp\mqty[\frac1N\sum_{n=1}^N(x_n-\mu_{ML})^2]=\frac1N\sum_{n=1}^N\Exp\mqty[(x_n-\mu_{ML})^2]\\
    &=\frac1N\sum_{n=1}^N\qty(\Var\mqty[x_n-\mu_{ML}]+(\Exp\mqty[x_n-\mu_{ML}])^2)=\frac{N-1}{N}\sigma^2.
\end{split}
$$


Therefore $\sigma^2_{ML}$ is biased, and the unbiased variance estimation is 

$$
    \tilde{\sigma}^2=\frac{1}{N-1}\sum_{n=1}^N(x_n-\mu_{ML})^2.
$$



## 


::: {#thm-bayesthm}
$$
f_{X\mid Y=y}(x)=\frac{f_{Y\mid X=x}(y\mid x)f_X(x)}{f_Y(y)}
$$
:::



::: {.callout-note}
$f_{X\mid Y}(x\mid y)$ is a pdf w.r.t $X$, not a pdf w.r.t $Y$.
:::




::: {.cell-output-stdout}
```
# Heading 1

## Heading 2

```
:::

